---
title: "Making nice looking crosstabulations using the tidyverse tools"
author: "Steve Simon"
source: new
date: "2025-09-17"
categories:
- Blog post
tags:
- R software
format: 
  html:
    embed-resources: true
editor: source
execute: 
  error: true
  echo: true
page_update: complete
---

Counts in a two by two table (a cross-tabulation), are tricky to display, especially when you want to include percentages. I want to illustrate some of the ways that you can use R to present these cross-tabulations. This page takes material from a [2002 webpage of mine][ref-simon-2002] on crosstabulation written from a more generic perspective.

[ref-simon-2002]: http://pmean.com/posts/displaying-percentages/

You might also want to look at [this vignette][ref-vignette] by Andri Signoreli written in 2021.

[ref-vignette]: https://stat.ethz.ch/CRAN/web/packages/DescTools/vignettes/TablesInR.pdf

There are plenty of other approaches out there that I will not try to cover here. These include

-   [Crosstable package][ref-Crosstable]
-   [data.table package][ref-data-table]
-   [DT package][ref-dt]
-   [flextable package][ref-flextable]
-   [gt package][ref-gt]
-   [janitor package][ref-janitor]

Some of these packages specialize in crosstabulations, but most provide them along with a range of other features.

[ref-Crosstable]: https://cran.r-project.org/web/packages/crosstable/vignettes/crosstable.html
[ref-data.table]: https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html
[ref-dt]: https://cran.r-project.org/web/packages/DT/refman/DT.html
[ref-flextable]: https://ardata-fr.github.io/flextable-book/
[ref-gt]: https://gt.rstudio.com/
[ref-janitor]: https://cran.r-project.org/web/packages/janitor/refman/janitor.html

<!---more--->

```{r}
#| label: setup
#| message: false
#| warning: false

library(glue)
library(gt)
library(tidyverse)
```

## Money and happiness

Does money buy happiness? Maybe, maybe not. Here are the first few rows of an artificial dataset where 200 hypothetical subjects were classified by income (rich or poor) and their mood (happy or miserable). You will see how to use R to display this data in a variety of different cross-tabulations.

```{r}
#| label: create-data
#| echo: false

# I am hiding the steps to create the data files used in this program because
# they are just a distraction from the main topic.

table_counts <- c(30, 10, 90, 70)
income <- c("rich", "rich", "poor", "poor")
mood <- c("happy", "miserable", "happy", "miserable")

mbh0 <- tibble(income, mood, table_counts)

# layout the data in long form, but a mostly random order

tibble(
    income=rep(income, table_counts-1), 
    mood=rep(mood, table_counts-1)) |>
  mutate(random_order=sample(1:196)) |>
  arrange(random_order) |>
  select(-random_order) -> mbh1

# I am mostly arranging the data in a random order, but I want to make sure 
# that all four combinations of rich/poor and happy/miserable are seen in the
# first four rows.

mbh1 <- bind_rows(mbh0[c(4, 1, 3, 2) , -3], mbh1) 
  
```

```{r}
#| label: long-form

mbh1 # "mbh" is short for "money buys happiness"
```

## Alternative presentation

Sometimes you will see this data presented in a summarized form. Here is what that might look like.

```{r}
#| label: summarized-form

mbh0
```


## Part 1, Crosstabulations

## Using the table function

This is the classic two by two crosstabulation.

```{r}
#| label: table

table(mbh1$income, mbh1$mood)
```

#### Comments on the code

Before the tidyverse came along, you might use the [table function][ref-table] to create a crosstabulation.

[ref-table]: https://stat.ethz.ch/R-manual/R-patched/RHOME/library/base/html/table.html

## Controlling the order of display

Now for this and all future tables, I want to show the results for rich income first and poor income second. There is no good reason for this other than that is how I arranged things in the 2002 webpage.

```{r}
#| label: factor

mbh1$income <- factor(mbh1$income, levels=c("rich", "poor"))
mbh0$income <- factor(mbh0$income, levels=c("rich", "poor"))
```

#### Comments on the code

The default in R is to arrange your string data in alphabetical order when displaying crosstabulations. You can override this default with the [factor function][ref-factor], part of base R. Use the levels argument to desigate the desired order.

[ref-factor]: https://stat.ethz.ch/R-manual/R-devel/library/base/html/factor.html

## Table after applying the factor function

Here is the table with the rows swapped.

```{r}
#| label: swap

table(mbh1$income, mbh1$mood)
```

## Turning the crosstabulation (table)

It is easy to turn the crosstabulation 90 degrees (swapping the rows and columns). Just place mood in the first argument and income in the second argument.

```{r}
#| label: turn

table(mbh1$mood, mbh1$income)
```

## Using the xtabs function

```{r}
#| label: xtabs-1

xtabs(formula= ~ income + mood, data=mbh1)
```

#### Comments on the code

Base R also has the [xtabs function][ref-xtabs]. This performs similarly to table, but has a few extra features. One of the more convenient features is the data argument which allows you to specify the data frame just once.

[ref-xtabs]: https://stat.ethz.ch/R-manual/R-devel/library/stats/html/xtabs.html

## Turning the crosstabulation (xtabs)

Swap the order to switch rows and columns.

```{r}
#| label: xtabs-2

xtabs(formula= ~ mood + income, data=mbh1)
```

## Creating a crosstabulation from an aggregated dataset

```{r}
#| label: xtabs-3

xtabs(table_counts~income+mood, data=mbh0)
```

#### Comments on the code

The xtabs function also allows you to create a table where you already have the counts. List the count or weight variable to the left of the tilde. What xtabs is doing here is just rearranging the four row by one column vector into a two row by two column matrix. You could have used the pivot_wider function here as well (see below for details on pivot_wider).

## Using the count function in dplyr

```{r}
#| label: count

mbh1 |>
  count(income, mood)
```

#### Comments on the code

I prefer using the [count function][ref-count] in dplyr instead of table or xtabs. It can mean a bit more work at times, but it is more flexible and can handle more complex settings. The count function produces the same results as table, but notice how it is laid out differently. The counts are in a single 4 by 1 column.

[ref-count]: https://dplyr.tidyverse.org/reference/count.html

## Converting from a column to a grid

```{r}
#| label: pivot-1

mbh1 |>
  count(income, mood) |>
  pivot_wider(names_from=mood, values_from=n)
```

#### Comments on the code

If you want to arrange this in a rectangular grid, you should use the [pivot_wider function][ref-pivot-wider], part of the tidyr/tidyverse library. 
This places income in the rows and mood in the columns.

[ref-pivot-wider]: https://tidyr.tidyverse.org/reference/pivot_wider.html

## Turning the crosstabulation (count/pivot_wider)

```{r}
#| label: pivot-2

mbh1 |>
  count(income, mood) |>
  pivot_wider(names_from=income, values_from=n)
```

#### Comments on the code

If you wanted the opposite, change the names_from argument to income.

## Part 2, adding totals to a table

When you use the table function or the xtabs function, your output is stored as a matrix. When you use the count function (and modify it with the pivot_wider function), your output is stored as a tibble. There are advantages and disadvantages to storing output in either format. The unfortunate result, however, is that further manipulations are handled differently for the different outputs.

## Adding row totals to a matrix created by table

```{r}
#| label: addmargins-2

table(mbh1$income, mbh1$mood) |>
  addmargins(margin=2)
```

#### Comments on the code

The [addmargins function][ref-addmargins] is part of base R and allows you to add row totals with the argument margin=2. 

[ref-addmargins]: https://stat.ethz.ch/R-manual/R-devel/library/stats/html/addmargins.html

## Adding column totals to a matrix created by table

You can add column totals with margin=1.

```{r}
#| label: row-margins

table(mbh1$income, mbh1$mood) |>
  addmargins(margin=1)
```

## Adding both row and column totals to a matrix created by table

You can add both row and column totals with margin=1:2.

```{r}
#| label: both-margins-1

table(mbh1$income, mbh1$mood) |>
  addmargins(margin=1:2)
```

## Adding row totals to a matrix created by xtabs

The addmargins function works equally well with the output from the xtabs function

```{r}
#| label: both-margins-2

xtabs(~ income + mood, data=mbh1) |>
  addmargins(margin="income")
```

## Adding row margins to a tibble created by count

Things work quite differently for the output from the count function. After pivoting, just add. Getting totals at the bottom of each column is not easy. In most real-world applications, you should be able to get by with just row margins.

```{r}
#| label: income-totals-1

mbh1 |>
  count(income, mood) |>
  group_by(income) |>
  mutate(total=sum(n)) 
```

If you want row totals after applying the pivot_wider function, just use a simple addition inside the mutate function.

```{r}
#| label: income-totals-2

mbh1 |>
  count(income, mood) |>
  pivot_wider(names_from=mood, values_from=n) |>
  mutate(total=happy+miserable) 
```

## Calculating proportions

There are multiple percentages that you can calculate and you need to plan this carefully. As a general rule, you usually (but not always) want to have the percentages represent a conditional probability. Also as a general rule, you want to probability to condition on a "cause" variable rather than on an "effect" variable.

It's not always obvious which variable is which. Exposure or intervention variables are usually causes. Demographic variables like race and sex are usually cause variables. Outcome measures are almost always effect variables. If you are unsure, try it both ways.

In this contrived example, consider income as cause and mood as an effect. Here is how to compute the probability of happy or miserable moods given income.

Again, the calculations depend on whether you are working with the matrix output of table and xtabs or the tibble output from count.

## Getting row proportions using a matrix created by xtabs

The [proportions function][ref-proportions] function takes a table object and computes row percents or column percents. Row percents add up to 100% in each row and usually (but not always) are what you want. The proportions function works easiest with output from xtabs because you can use names rather than numbers. With table, you use 1 for row percents and 2 for column percents.

```{r}
#| label: conditional-1

xtabs( ~ income + mood, data=mbh1) |>
  proportions(margin="income")
```

[ref-proportions]: https://stat.ethz.ch/R-manual/R-devel/library/base/html/proportions.html

## Getting row proportions using a tibble created by count

When you want row percents from a tibble created by the count function, you add a few steps before you pivot wider. 

```{r}
#| label: conditional-2

mbh1 |>
  count(income, mood) |>
  group_by(income) |>
  mutate(total=sum(n)) |>
  mutate(prop=n/total) |>
  select(-n, -total) |>
  pivot_wider(names_from=mood, values_from=prop)
```

## Running statistical tests

You may want to use various statistical tests on your crosstabulation. Some possible choices might be

-   [chisq.test][ref-chisq]
-   [fisher.test][ref-fisher]
-   [prop.test][ref-prop]
-   [riskratio][ref-riskratio] (from the epitools library)

[ref-chisq]: https://stat.ethz.ch/R-manual/R-devel/library/stats/html/chisq.test.html
[ref-fisher]: https://stat.ethz.ch/R-manual/R-devel/library/stats/html/fisher.test.html
[ref-prop]: https://stat.ethz.ch/R-manual/R-devel/library/stats/html/prop.test.html
[ref-riskratio]: https://cran.r-project.org/web/packages/epitools/refman/epitools.html#riskratio

All of these functions will work well with a matrix created by table or xtabs. They will not work well with the tibble created by count.

If you use logistic regression (the [glm function][ref-glm] with the argument family=binomial), you are best off using the raw data rather than a summary with table, xtabs, or count.

[ref-glm]: https://stat.ethz.ch/R-manual/R-devel/library/stats/html/glm.html

## Making it look nice

One advantage to using the tibble output from count is that it allows you a lot of flexibility to make the output look a little bit closer (or a lot closer) to publication quality. Here is a simple example.

```{r}
#| label: layout

mbh1 |>
  count(income, mood) |>
  group_by(income) |>
  mutate(total=sum(n)) |>
  mutate(pct=round(100*n/total)) |>
  mutate(pct=glue("{n}/{total} = {pct}%")) |>
  select(-n, -total) |>
  pivot_wider(names_from=mood, values_from=pct)
```

#### Comments on the code

The [glue function][ref-glue-1], part of the [glue library][ref-glue-2], can produce a layout that is a bit closer to publication quality. There are several possible layouts, but here is one that I like.

[ref-glue-1]: https://glue.tidyverse.org/reference/glue.html
[ref-glue-2]: https://glue.tidyverse.org/

